---
title: "Reproducible Research: Peer Assessment 1"
author: "Tamer Salem"
date: "March 9, 2020"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.path="figure/activity")
```
```{r echo=TRUE, cache=TRUE}
# Operate in current directory

```

## Loading and preprocessing the data

```{r echo=TRUE, cache=TRUE}
activity.data.raw <- read.csv(file="activity.csv", header=TRUE, sep=",")
print(paste0("Number Rows (raw data)=", nrow(activity.data.raw)))
print(head(activity.data.raw))
```

The original data has many incomplete cases, i.e., those with NA values.  Reduce the set down to those with complete data.
```{r echo=TRUE, cache=TRUE}
activity.data.tidy <- activity.data.raw[complete.cases(activity.data.raw),]
print(paste0("Number Rows (tidy data)=", nrow(activity.data.tidy)))
print(head(activity.data.tidy))
```

## What is mean total number of steps taken per day?

Ignore the incomplete data - use the tidy set.  

First get the unique days.  Then, count the number of step taken each day.
```{r echo=TRUE, cache=TRUE}
unique.days <- unique(activity.data.tidy[,2])
steps.per.day <- integer(length=length(unique.days))
i <- 1
for( day in unique.days ) {
  day.values <- activity.data.tidy[activity.data.tidy$date==day,]
  steps.per.day[i] <- sum(day.values[,1])
  i <- i+1
}
```
Create a histogram of the number of steps per day.
```{r echo=TRUE, cache=TRUE}
hist(steps.per.day, main="Steps per Day (Tidy Data)", xlab="Number of Steps per Day", ylab="Number of Days", breaks=12,
     xlim=c(0,25000), ylim=c(0,20))
```
Mean and median of steps per day:
```{r echo=TRUE, cache=TRUE}
print(paste0("Mean (Steps per day): ", mean(steps.per.day)))
print(paste0("Median (Steps per day): ", median(steps.per.day)))
```

## What is the average daily activity pattern?

Extract the intervals and calculate the mean of the daily number of steps in each of those intervals.
```{r echo=TRUE, cache=TRUE}
unique.intervals <- unique(activity.data.tidy[,3])
mean.steps.per.interval <- integer(length=length(unique.intervals))
i <- 1
for( interval in unique.intervals ) {
  interval.values <- activity.data.tidy[activity.data.tidy$interval==interval,]
  mean.steps.per.interval[i] <- mean(interval.values[,1])
  i <- i+1
}
df <- data.frame(intervals=unique.intervals, mean_steps=mean.steps.per.interval)
plot(x=df$intervals, y=df$mean_steps, type="l", main="Average Daily Activity Pattern (Tidy Data)", xlab="Interval of Day",
     ylab="Mean Number of Steps")
```


## Imputing missing values

```{r echo=TRUE, cache=TRUE}
number.NAs <- nrow(activity.data.raw) - nrow(activity.data.tidy)
```

There are `r nrow(activity.data.raw)` total rows in the original data.  There are `r number.NAs` missing values (rows with NAs) in the original data set.  A reasonable strategy for imputing the 
values for the missing steps is to use the mean number of steps for the interval whose value is missing.  That
value can be found from the data frame *df*.  Using that strategy, synthesize a complete data set.
```{r echo=TRUE, cache=TRUE}
activity.synthesized.data <- activity.data.raw
for( i in 1:nrow(activity.synthesized.data) ) {
  if( is.na(activity.synthesized.data[i,]$steps) ) {
    activity.synthesized.data[i,]$steps <- df[df$intervals==activity.synthesized.data[i,]$interval,]$mean_steps
  }
}
# Double check that there are no NAs left
print(summary(complete.cases(activity.synthesized.data)))
```
Now, recreate the histogram of the data and report the mean and median values on the synthesized data with imputed values.  
```{r echo=TRUE, cache=TRUE}
unique.days.imputed <- unique(activity.synthesized.data[,2])
steps.per.day.imputed <- integer(length=length(unique.days.imputed))
i <- 1
for( day in unique.days.imputed ) {
  day.values.imputed <- activity.synthesized.data[activity.synthesized.data$date==day,]
  steps.per.day.imputed[i] <- sum(day.values.imputed[,1])
  i <- i+1
}
hist(steps.per.day.imputed, main="Steps per Day (Imputed Data)", xlab="Number of Steps per Day", ylab="Number of Days", breaks=12,
     xlim=c(0,25000), ylim=c(0,25))
```

The mean and mean number of (imputed) steps per day:
```{r echo=TRUE}
print(mean(steps.per.day.imputed))
print(median(steps.per.day.imputed))
```
To better see the effect of imputing the data, use a data frame
```{r echo=TRUE, cache=TRUE}
effect.df <- data.frame(mean=c(mean(steps.per.day),mean(steps.per.day.imputed)),
                        median=c(median(steps.per.day),median(steps.per.day.imputed)))
row.names(effect.df) <- c("Tidy", "Imputed")
print(effect.df)
```
The fact that the mean values are equivalent should be expected since we are assigning the missing values as the mean
values.  We note that the difference in the mean and median values of the tidy set is small, so it is not surprising
that the difference in the median values between the tidy and synthesized (imputed) data is also very small.  It is also
interesting that the median vale for the synthesized set is the same as the mean value - likely the effect of assigning
so many mean values back into the synthesized set.

## Are there differences in activity patterns between weekdays and weekends?

Add a factor variable (as a column) to the synthesized dataset to distinguish between weekdays and weekends.
```{r echo=TRUE, cache=TRUE}
days.of.week <- weekdays(as.Date(activity.synthesized.data$date))
day.factor <- lapply(days.of.week, function(x) if(x=="Saturday" || x=="Sunday") "weekend" else "weekday")
day.factor <- as.factor(unlist(day.factor))
cnames <- colnames(activity.synthesized.data)
sap.activity.data <- cbind(activity.synthesized.data, day.factor)
colnames(sap.activity.data) <- c(cnames, "weekday_or_weekend")
print(head(sap.activity.data))
```

Create a panel plot (time series) for weekday and weekend values.

```{r echo=TRUE, cache=TRUE}
wk.end <- sap.activity.data[sap.activity.data$weekday_or_weekend=="weekend",]
wk.day <- sap.activity.data[sap.activity.data$weekday_or_weekend=="weekday",]
unique.intervals <- unique(wk.end[,3])
mean.steps.per.interval <- integer(length=length(unique.intervals))
i <- 1
for( interval in unique.intervals ) {
  interval.values <- wk.end[wk.end$interval==interval,]
  mean.steps.per.interval[i] <- mean(interval.values[,1])
  i <- i+1
}
wk.end.df <- data.frame(intervals=unique.intervals, mean_steps=mean.steps.per.interval)
unique.intervals <- unique(wk.day[,3])
mean.steps.per.interval <- integer(length=length(unique.intervals))
i <- 1
for( interval in unique.intervals ) {
  interval.values <- wk.day[wk.day$interval==interval,]
  mean.steps.per.interval[i] <- mean(interval.values[,1])
  i <- i+1
}
wk.day.df <- data.frame(intervals=unique.intervals, mean_steps=mean.steps.per.interval)
layout(matrix(1:2,ncol=1),widths=1,heights=c(2,2),respect=FALSE)
par(mar = rep(0, 4), oma=c(4, 4, 4, 2), las=1)
plot(x=wk.day.df$intervals, y=wk.day.df$mean_steps, type="l", ann=FALSE, xaxt='n', col=2)
legend(1900,170,legend=c("Weekday"), yjust=0)
plot(x=wk.end.df$intervals, y=wk.end.df$mean_steps, type="l", ann=FALSE, col=3)
legend(1900,120,legend=c("Weekend"), yjust=0)
mtext("Interval of Day", side=1, line=3, outer=TRUE)
mtext("Mean Number of Steps", side=2, line=3, outer=TRUE, las=0)
mtext("Average Activity Pattern (Synthesized Data)", side=3, line=1, outer=TRUE, cex=2, font=1)

```
